The Structure of $k$-Planar Graphs

### 2019-12-05 k-planar.tex
Nearly done

\documentclass[a4paper,UKenglish,cleveref, autoref]{lipics-v2019}

@@

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,wrapfig}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{amsfonts}
\usepackage{amsthm,mathtools}
\usepackage{pat}
\usepackage{paralist}
\usepackage{stmaryrd}
\usepackage{thm-restate}

@@

\allowdisplaybreaks
%\sloppy
%%%%%%%%%%
\makeatletter
\def\NAT@spacechar{~}
\makeatother
%%%%%%%%%%
\crefname{lem}{Lemma}{Lemmas}
\crefname{thm}{Theorem}{Theorems}
\crefname{cor}{Corollary}{Corollaries}
\crefname{conj}{Conjecture}{Conjectures}
\crefname{obs}{Observation}{Observations}
\crefname{prop}{Proposition}{Propositions}
\crefname{clm}{Claim}{Claims}
\crefname{openproblem}{Open Problem}{Open Problems}
\crefformat{equation}{(#2#1#3)}
\Crefformat{equation}{Equation #2(#1)#3}

@@

\author{Vida Dujmović}{School of Computer Science and Electrical Engineering, University of Ottawa}{vida.dujmovic@uottawa.ca}{[0000-0003-0471-4118]}{Research  supported by NSERC and the Ontario Ministry of Research and Innovation.}
% , Ottawa, Canada (\texttt{vida.dujmovic@uottawa.ca}).
%                 Research  supported by NSERC and the Ontario Ministry of Research and Innovation.},\,\,
\author{Pat Morin}{School of Computer Science, Carleton University}{morin@scs.carleton.ca}{[0000-0003-0471-4118]}{Research supported by NSERC.}
% , Ottawa, Canada (\texttt{morin@scs.carleton.ca}).                 Research  supported by NSERC and the Ontario Ministry of Research and Innovation.},\,\, and
\author{David R. Wood}{School of Mathematics, Monash University}{david.wood@monash.edu}{[0000-0003-0471-4118]}{Research supported by the Australian Research Council.}
% , Melbourne, Australia (\texttt{david.wood@monash.edu}). Research supported by the Australian Research Council.}
% }

@@

\Copyright{Vida Dujmović, Pat Morin, and David R. Wood}

\begin{document}
% \begin{titlepage}
\maketitle

\begin{abstract}
Dujmovi\'c~et~al.~(FOCS 2019) recently proved that every planar graph is a subgraph of the strong product of a graph of bounded treewidth and a path. This tool has been used to solve longstanding problems on queue layouts, non-repetitive colouring, $p$-centered colouring, and implicit graph encoding. We extend this result for $k$-planar graphs, where a graph is \emph{$k$-planar} if it has a drawing in the plane in which each edge is involved in at most $k$ crossings. In particular, we prove that every $k$-planar graph is a subgraph of the strong product of a graph of treewidth $O(k^5)$ and a path. This is the first result of this type for a non-minor-closed class of graphs. It implies, amongst other results, that $k$-planar graphs have non-repetitive chromatic number upper-bounded by a function of $k$. All these results generalise for drawings of graphs on arbitrary surfaces. In fact, we work in a much more general setting based on so-called shortcut systems that are of independent interest. 
\end{abstract}
% \end{titlepage}

@@

Although $k$-planar graphs are the most high-profile target for a generalization of \cref{PlanarBasic}, we actually prove a substantially stronger result than \cref{kPlanarBasic} using the following definition. A non-empty set $\mathcal{P}$ of non-trivial paths in a graph $G$ is a \emph{$(k,d)$-shortcut system} (for $G$) if:
\begin{compactitem}
\item every path in $\mathcal{P}$ has length at most $k$,\footnote{A path of length $k$ consists of $k$ edges and $k+1$ vertices.  A path is \emph{trivial} if it has length 0 and \emph{non-trivial} otherwise.} and
\item for every $v\in V(G)$, the number of paths in $\mathcal{P}$ that use $v$ as an internal vertex is at most $d$.
\end{compactitem} 
Each path $P\in\mathcal{P}$ is called a \emph{shortcut}; if $P$ has endpoints $v$ and $w$ then it is a \emph{$vw$-shortcut}. Given a graph $G$ and a $(k,d)$-shortcut system $\mathcal{P}$ for $G$, let $G^{\mathcal{P}}$ denote the supergraph of $G$ obtained by adding the edge $vw$ for each $vw$-shortcut in $\mathcal{P}$.

@@

We prove the following theorem that shows if a graph $G$ is a subgraph of $H\boxtimes P$ and $\PP$ is a shortcut system for $G$, then $G^{\mathcal{P}}$ is a subgraph of $J\boxtimes P$, where the treewidth of $J$ is bounded by a function of the treewidth of $H$.

@@

\cref{ShortcutProduct} leads to several other results of interest. Here is one example. The \emph{$k$-th power} of a graph $G$ is the graph $G^k$ with vertex set $V(G^k):=V(G)$, where $vw\in E(G^k)$ if and only if $\dist_G(v,w)\leq k$.\footnote{For a graph $G$ and two vertices $v,w\in V(G)$, $\dist_G(v,w)$  denotes the length of a shortest path, in $G$, with endpoints $v$ and $w$.  We define $\dist_G(v,w):=\infty$ if $v$ and $w$ are in different connected components of $G$.} If $G$ has maximum degree $\Delta$, then $G^k = G^\PP$ for some $(k,2k\Delta^{k})$-shortcut system $\PP$; see \cref{PowerShortcut}. \cref{kPlanarBasic,ShortcutProduct} then imply:

@@

This section defines concepts and results from the literature that will be important for our work.

@@

\citet{dujmovic.joret.ea:planar} introduced the study of partitions with bounded layered width such that the quotient has some additional desirable property, like small treewidth. 
% Say that a graph $G$ admits a $(t,\ell)$-partition if $V(G)$ has a partition $\mathcal{P}$ of layered width at most $\ell$ and the treewidth of $H:=G/\mathcal{P}$ is at most $t$. 
They defined a class $\mathcal{G}$ of graphs to \emph{admit bounded layered partitions} if there exist $t,\ell\in\mathbb{N}$ such that every graph $G\in \mathcal{G}$ has an $H$-partition of layered width at most $\ell$ for some graph $H=H(G)$ of treewidth at most $t$.

These definitions relate to strong products as follows.

@@

\begin{thm}[\citep{dujmovic.joret.ea:planar}]
\label{SurfacePartition}
Every graph of Euler genus $g$ is a subgraph of:
\begin{compactenum}[(a)]
\item  $H \boxtimes P \boxtimes K_{\max\{2g,1\}}$ for some graph $H$ of treewidth at most $9$ and for some path $P$;
\item  $H \boxtimes P \boxtimes K_{\max\{2g,3\}}$ for some graph $H$ of treewidth at most $4$ and for some path $P$.
%\item $(K_{2g} + H )  \boxtimes P$ for some graph $H$ of treewidth at most $8$  and some path $P$. 
\end{compactenum}
Equivalently, every graph of Euler genus $g$ has:
\begin{compactenum}[(a)]
\item an $H$-partition with layered width at most $\max\{2g,1\}$ such that $H$ has treewidth at most $9$;
\item an $H$-partition with layered width at most $\max\{2g,3\}$ such that $H$ has treewidth at most $4$.
%\item an $H$-partition with layered width at most $1$ such that $H$ has treewidth at most $2g+8$. 
\end{compactenum}
%PM: Keep 'em all: \textcolor{red}{eliminate some of these results? just keep (b)? }
\end{thm}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Shortcut Systems}
\label{Structure}

@@

For a tree $T$ rooted at some node $x_0\in V(T)$, we we say that a node $a\in V(T)$ is a \emph{$T$-ancestor} of $x\in V(T)$ (and $x$ is a \emph{$T$-descendant} of $a$) if $a$ is a vertex of the path, in $T$, from $x_0$ to $x$.  Note that each node $x\in v(T)$ is a $T$-ancestor and $T$-descendant of itself.  We say that a $T$-ancestor $a\in V(T)$ of $x\in V(T)$ is a \emph{strict} $T$-ancestor of $x$ if $a\neq x$. 
The \emph{$T$-depth} of a node $x\in V(T)$ is the length of the path, in $T$, from $x_0$ to $x$.  For each node $x\in V(T)$, define
\[T_x := T[\{y\in V(T):\mbox{$x$ is a $T$-ancestor of $y$}\}] \]
to be the maximal subtree of $T$ rooted at $x$.

We begin with a fairly standard lemma about normalised tree decompositions.

\begin{lem}\lemlabel{nice-decomposition}
  For every graph $H$ of treewidth $t$, there is a rooted tree $T$ with $V(T)=V(H)$ and a $T$-decomposition $(B_x:x\in V(T))$ of $H$ with width $t$ that has following additional properties:  
  \begin{compactenum}[(T1)]
    % \item\tlabel{rooted}\tlabel{first} $T$ is rooted at some node $x_0\in V(T)$ with $B_{x_0}=\emptyset$.
    % \item\tlabel{node-set} $V(T)= V(H)$.
    % \item\tlabel{diff} For each edge $xy\in E(T)$, $|B_x\ominus B_y|\le 1$.
    \item\tlabel{subtree-root} For each node $x\in V(H)$, the subtree $T[x]:=T[\{y\in V(T):x\in B_y\}]$ is rooted at $x$.
    \item\tlabel{ancestor-edge}\tlabel{last} For each edge $xy\in E(H)$, one of $x$ or $y$ is a $T$-ancestor of the other.
  \end{compactenum}
\end{lem}

@@

Before proving \cref{generalized-tripod} we point out more properties that are immediately implied by it:

@@

\begin{proof}[Proof of \cref{generalized-tripod}]
  Property~\yref{thickness} follows immediately from the fact that $V(T)=V(H)$
  and the fact that $\mathcal{Y}$ has layered width at most $\ell$ with respect to $\mathcal{L}$.  Property \yref{separator} is immediate from the definitions of $F_x$ and $N_x$.  In particular, $(N_x,V(G)\setminus V_x)$ is a separation of $G$ with $F_x=N_x\cap(V(G)\setminus V_x)$.

@@

\begin{proof}
Apply \cref{generalized-tripod} to $G$ and let $\mathcal{L}$, $\mathcal{Y}$, $\mathcal{T}$, $T$, $Y_x$, $V_x$, $F_x$, and $N_x$ be defined as in \cref{generalized-tripod}.

For a node $x\in V(T)$, we say that a shortcut $P\in\mathcal{P}$ \emph{crosses} $x$ if $Y_x$ contains an internal vertex of $P$, that is, $P=(v_0,\ldots,v_r)$ and $\{v_1,\ldots,v_{r-1}\}\cap Y_x\neq\emptyset$.  We say that a vertex $v\in V(G)$ \emph{participates} in $x$ if $v\in Y_x$ or $\mathcal{P}$ contains a shortcut $P$ with $v\in V(P)$ and $P$ crosses $x$. We let $X_v$ denote the set of nodes $x\in V(T)$ such that $v$ participates in $x$.

@@

\begin{proof}
  Let 
  \[
     Z := \{v\} \cup \{\{v_1,\ldots,v_{r-1}\}:\mbox{$(v_0,\ldots,v_r)\in\mathcal{P}$, $v\in \{v_0,\ldots,v_r\}$}\}
  \]
  Clearly $G[Z]$ is connected because $Z$ is the union of (vertex sets of) paths in $G$, each of which contains $v$.

@@

  Let $Z_H:=\{x\in V(H): Z\cap Y_x\neq\emptyset\}$.  The connectivity of $G[Z]$ implies that $H[X_H]$ is connected.  
  %Property~\tref{ancestor-edge} implies that, for every $xy\in E(H)$, one of $x$ or $y$ is a $T$-ancestor of the other. 
  Choose $a(v)\in X_H$ to be any member of $X_H$ that does not have a strict $T$-ancestor in $X_H$.  Transitivity of the $T$-ancestor relationship, \tref{ancestor-edge}, and connectivity of $H[X_H]$ implies that $a(v)$ is a $T$-ancestor of every node $x\in X_H$, as required.
\end{proof}

For each $x\in V(T)$, we define the \emph{separator}
\[  
     S_x := \{v\in V(G): a(v)= x\} \enspace .
\]
Observe that $\mathcal{S}:=(S_x : x\in V(T))$ is a partition of $V(G)$.  We let $J:=G^\mathcal{P}/\mathcal{S}$ denote the resulting quotient graph and we let $V(J)\subseteq V(T)$ in the obvious way, so that each $x\in V(J)$ is the vertex obtained by contracting $S_x$ in $G^{\mathcal{P}}$. (Nodes $x\in V(T)$ with $S_x=\emptyset$ do not contribute a vertex to $J$.)

@@

%%%%%%%%%%%%%%%%%%%
\section{\boldmath $k$-Planar Graphs}
\label{sec-k-planar}

@@

We can reduce the layered width of the $H$-partition of $G$ from $O(k^3)$ to $O(k^2)$ by observing that the dummy vertices in $V(G_0)\setminus V(G)$ do not contribute to the layered width of this partition.  In this setting, the proof of \cref{clm:general-width} is simpler since each vertex $w\in Y_x$ contributes at most two vertices to $V_i\cap Y_x$.  More precisely, each path $P\in\mathcal{P}$ containing an internal (dummy) vertex $w\in Y_x\cap (V_{i-j}\cup V_{i+j})$ contributes: (i)~at most two vertices to $S_x\cap V_i$ for $j\in\{0,\ldots,\floor{(k+1)/2}\}$; (ii)~at most one vertex to $S_x\cap V_j$ for $j\in\{\floor{(k+1)/2}+1,\ldots,k+1\}$; or (iii)~no vertices to $S_x\cap V_j$ for $j > k+1$.
Redoing the calculation at the end of the proof of \cref{clm:general-width} then yields 
\begin{align*}
    |S_x\cap Y_i| \le d\ell\left( 
       2 
       + 4\left\lfloor\frac{k+1}{2}\right\rfloor 
       + 2\left\lceil\frac{k+1}{2}\right\rceil
       \right)
       & =
       d\ell\left(
         2 + 2(k+1) + 2\left\lfloor\frac{k+1}{2}\right\rfloor
       \right) \\
       & \le
       d\ell(3k+5) 
      = 18k+30 \enspace .
\end{align*}    
With this change, the layered width of the partition given by \cref{modern-major-general} becomes $(18k+30)(k+1)=18k^2+48k+30$.  This establishes \cref{k-planar}.
% Note to self-and-others. In our original SODA submission we had 18k^2+30k here because we forgot that the layer compression step has compress k+1 consecutive layers, not just k consecutive layers.

Exactly the same approach using \cref{SurfacePartition}(b) instead of \cref{PlanarPartition}(b) establishes 
\cref{g-k-planar} for $(g,k)$-planar graphs.

\subsection{\boldmath  $1$-Planar Graphs}
\label{sec-1-planar}

>>>>

 \documentclass[a4paper,UKenglish,cleveref,autoref]{lipics-v2019}

@@

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{amsfonts}
\usepackage{amsthm,mathtools}
\usepackage{pat}
\usepackage{paralist}
% \usepackage{stmaryrd}
\usepackage{thm-restate}

@@

% \allowdisplaybreaks
% %\sloppy
% %%%%%%%%%%
% \makeatletter
% \def\NAT@spacechar{~}
% \makeatother
%%%%%%%%%%
\crefname{lem}{Lemma}{Lemmas}
\crefname{thm}{Theorem}{Theorems}
\crefname{cor}{Corollary}{Corollaries}
\crefname{conj}{Conjecture}{Conjectures}
\crefname{obs}{Observation}{Observations}
\crefname{prop}{Proposition}{Propositions}
\crefname{clm}{Claim}{Claims}
\crefname{openproblem}{Open Problem}{Open Problems}
\crefformat{equation}{(#2#1#3)}
\Crefformat{equation}{Equation #2(#1)#3}

@@

\author{Vida Dujmović}{School of Computer Science and Electrical Engineering, University of Ottawa}{vida.dujmovic@uottawa.ca}{}{Research  supported by NSERC.}
% , Ottawa, Canada (\texttt{vida.dujmovic@uottawa.ca}).
%                 Research  supported by NSERC and the Ontario Ministry of Research and Innovation.},\,\,
\author{Pat Morin}{School of Computer Science, Carleton University}{morin@scs.carleton.ca}{[0000-0003-0471-4118]}{Research supported by NSERC.}
% , Ottawa, Canada (\texttt{morin@scs.carleton.ca}).                 Research  supported by NSERC and the Ontario Ministry of Research and Innovation.},\,\, and
\author{David R. Wood}{School of Mathematics, Monash University}{david.wood@monash.edu}{}{Research supported by the Australian Research Council.}
% , Melbourne, Australia (\texttt{david.wood@monash.edu}). Research supported by the Australian Research Council.}
% }

@@

\Copyright{Vida Dujmović, Pat Morin, and David R. Wood}
\ccsdesc{Mathematics of computing~Graph theory}
\ccsdesc{Mathematics of computing~Extremal graph theory}
\ccsdesc{Human-centered computing~Graph drawings}
\relatedversion{This paper has been submitted to Discrete \&
Computational Geometry. The current status, as of December 4th, 2019, is ``under review''.  A full version of this paper is available at
                 \url{https://arxiv.org/abs/1907.05168}.}
\funding{This work was partly funded by NSERC and MRI.}

\begin{document}

\maketitle
\begin{abstract}
Dujmovi\'c~et~al.~(FOCS 2019) recently proved that every planar graph is a subgraph of the strong product of a graph of bounded treewidth and a path. This tool has been used to solve longstanding problems on queue layouts, non-repetitive colouring, $p$-centered colouring, and implicit graph encoding. We extend this result for $k$-planar graphs, where a graph is \emph{$k$-planar} if it has a drawing in the plane in which each edge is involved in at most $k$ crossings. In particular, we prove that every $k$-planar graph is a subgraph of the strong product of a graph of treewidth $O(k^5)$ and a path. This is the first result of this type for a non-minor-closed class of graphs. It implies, amongst other results, that $k$-planar graphs have non-repetitive chromatic number upper-bounded by a function of $k$. All these results generalise for drawings of graphs on arbitrary surfaces. In fact, we work in a much more general setting based on so-called shortcut systems that are of independent interest. 
\end{abstract}
% \end{titlepage}

@@

Although $k$-planar graphs are the most high-profile target for a generalization of \cref{PlanarBasic}, we actually prove a substantially stronger result than \cref{kPlanarBasic} using the following definition. A non-empty set $\mathcal{P}$ of non-trivial paths in a graph $G$ is a \emph{$(k,d)$-shortcut system} (for $G$) if:
\begin{compactitem}
\item every path in $\mathcal{P}$ has length at most $k$,\footnote{A path of length $k$ consists of $k$ edges and $k+1$ vertices.  A path is \emph{trivial} if it has length 0 and \emph{non-trivial} otherwise.} and
\item for every $v\in V(G)$, the number of paths in $\mathcal{P}$ with $v$ as an internal vertex is at most $d$.
\end{compactitem} 
Each path $P\in\mathcal{P}$ is called a \emph{shortcut}; if $P$ has endpoints $v$ and $w$ then it is a \emph{$vw$-shortcut}. Given a graph $G$ and a $(k,d)$-shortcut system $\mathcal{P}$ for $G$, let $G^{\mathcal{P}}$ denote the supergraph of $G$ obtained by adding the edge $vw$ for each $vw$-shortcut in $\mathcal{P}$.

@@

We prove the following theorem that shows if a graph $G$ is a subgraph of $H\boxtimes P$ and $\PP$ is a shortcut system for $G$, then $G^{\mathcal{P}}$ is a subgraph of $J\boxtimes P$, where the treewidth of $J$ is bounded by a function of $k$, $d$, and the treewidth of $H$.

@@

\cref{ShortcutProduct} leads to several other results. Here is one example. The \emph{$k$-th power} of a graph $G$ is the graph $G^k$ with vertex set $V(G^k):=V(G)$, where $vw\in E(G^k)$ if and only if $\dist_G(v,w)\leq k$.\footnote{For a graph $G$ and two vertices $v,w\in V(G)$, $\dist_G(v,w)$  denotes the length of a shortest path, in $G$, with endpoints $v$ and $w$.  We define $\dist_G(v,w):=\infty$ if $v$ and $w$ are in different connected components of $G$.} If $G$ has maximum degree $\Delta$, then $G^k = G^\PP$ for some $(k,2k\Delta^{k})$-shortcut system $\PP$; see \cref{PowerShortcut}. \cref{kPlanarBasic,ShortcutProduct} then imply:

@@

This section defines concepts and results from the literature that are used in our work.

@@

\citet{dujmovic.joret.ea:planar} introduced the study of partitions with bounded layered width such that the quotient has some additional desirable property, like small treewidth. 
% Say that a graph $G$ admits a $(t,\ell)$-partition if $V(G)$ has a partition $\mathcal{P}$ of layered width at most $\ell$ and the treewidth of $H:=G/\mathcal{P}$ is at most $t$. 
They defined a class $\mathcal{G}$ of graphs to \emph{admit bounded layered partitions} if there exist $t,\ell\in\mathbb{N}$ such that every graph $G\in \mathcal{G}$ has an $H$-partition of layered width at most $\ell$ for some graph $H=H(G)$ of treewidth at most $t$. These definitions relate to strong products as follows.

@@

\begin{thm}[\citep{dujmovic.joret.ea:planar}]
\label{SurfacePartition}
Every graph of Euler genus $g$ is a subgraph of:
\begin{compactenum}[(a)]
\item  $H \boxtimes P \boxtimes K_{\max\{2g,1\}}$ for some graph $H$ of treewidth at most $9$ and for some path $P$;
\item  $H \boxtimes P \boxtimes K_{\max\{2g,3\}}$ for some graph $H$ of treewidth at most $4$ and for some path $P$.
%\item $(K_{2g} + H )  \boxtimes P$ for some graph $H$ of treewidth at most $8$  and some path $P$. 
\end{compactenum}
Equivalently, every graph of Euler genus $g$ has:
\begin{compactenum}[(a)]
\item an $H$-partition with layered width at most $\max\{2g,1\}$ where $H$ has treewidth at most $9$;
\item an $H$-partition with layered width at most $\max\{2g,3\}$ where $H$ has treewidth at most $4$.
%\item an $H$-partition with layered width at most $1$ such that $H$ has treewidth at most $2g+8$. 
\end{compactenum}
%PM: Keep 'em all: \textcolor{red}{eliminate some of these results? just keep (b)? }
\end{thm}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Shortcut Systems: Proof of \cref{modern-major-general}}
\label{Structure}

@@

For a tree $T$ rooted at some node $x_0\in V(T)$, we say that a node $a\in V(T)$ is a \emph{$T$-ancestor} of $x\in V(T)$ (and $x$ is a \emph{$T$-descendant} of $a$) if $a$ is a vertex of the path, in $T$, from $x_0$ to $x$.  Note that each node $x\in V(T)$ is a $T$-ancestor and $T$-descendant of itself.  We say that a $T$-ancestor $a\in V(T)$ of $x\in V(T)$ is a \emph{strict} $T$-ancestor of $x$ if $a\neq x$. 
The \emph{$T$-depth} of a node $x\in V(T)$ is the length of the path, in $T$, from $x_0$ to $x$.  For each node $x\in V(T)$, define $T_x := T[\{y\in V(T):\mbox{$x$ is a $T$-ancestor of $y$}\}]$ to be the maximal subtree of $T$ rooted at $x$.

We begin with a fairly standard trick of working with a normalised tree decomposition.

\begin{lem}\lemlabel{nice-decomposition}
  For every graph $H$ of treewidth $t$, there is a rooted tree $T$ with $V(T)=V(H)$ and a $T$-decomposition $(B_x:x\in V(T))$ of $H$ of width at most $t$ that has following additional properties:  
  \begin{compactenum}[(T1)]
    % \item\tlabel{rooted}\tlabel{first} $T$ is rooted at some node $x_0\in V(T)$ with $B_{x_0}=\emptyset$.
    % \item\tlabel{node-set} $V(T)= V(H)$.
    % \item\tlabel{diff} For each edge $xy\in E(T)$, $|B_x\ominus B_y|\le 1$.
    \item\tlabel{subtree-root} For each node $x\in V(H)$, the subtree $T[x]:=T[\{y\in V(T):x\in B_y\}]$ is rooted at $x$.
    \item\tlabel{ancestor-edge}\tlabel{last} For each edge $xy\in E(H)$, one of $x$ or $y$ is a $T$-ancestor of the other.
  \end{compactenum}
\end{lem}

@@

We first point out more properties that are immediately implied by \cref{generalized-tripod}:

@@

\begin{proof}[Proof of \cref{generalized-tripod}]
  Property~\yref{thickness} follows immediately from the fact that $V(T)=V(H)$ (a condition ensured by \lemref{nice-decomposition}) and the fact that $\mathcal{Y}$ has layered width at most $\ell$ with respect to $\mathcal{L}$.  Property \yref{separator} is immediate from the definitions of $F_x$ and $N_x$.  In particular, $(N_x,V(G)\setminus V_x)$ is a separation of $G$ with $F_x=N_x\cap(V(G)\setminus V_x)$.

@@

\begin{proof}
Apply \cref{generalized-tripod} to $G$ and let $\mathcal{L}$, $\mathcal{Y}$, $\mathcal{T}$, $T$, $Y_x$, $V_x$, $F_x$, and $N_x$ be defined as in \cref{generalized-tripod} (so that $(\mathcal{L},\mathcal{Y})$ is an $H$-decomposition of $G$ of layered width $\ell$ where $H$ has treewidth $t$.)

For a node $x\in V(T)$, we say that a shortcut $P\in\mathcal{P}$ \emph{crosses} $x$ if $Y_x$ contains an internal vertex of $P$, that is, $P=(v_0,\ldots,v_r)$ and $\{v_1,\ldots,v_{r-1}\}\cap Y_x\neq\emptyset$.  We say that a vertex $v\in V(G)$ \emph{participates} in $x$ if $v\in Y_x$ or $\mathcal{P}$ contains a shortcut $P$ with $v\in V(P)$ and $P$ crosses $x$. Let $X_v$ denote the set of nodes $x\in V(T)$ such that $v$ participates in $x$.

@@

\begin{proof}
  Let  $Z := \{v\} \cup \{\{v_1,\ldots,v_{r-1}\}:(v_0,\ldots,v_r)\in\mathcal{P}$, $v\in \{v_0,\ldots,v_r\}\}$
  Clearly $G[Z]$ is connected because $Z$ is the union of (vertex sets of) paths in $G$, each of which contains $v$.

@@

  Let $X_H:=\{x\in V(H): Z\cap Y_x\neq\emptyset\}$.  The connectivity of $G[Z]$ implies that $H[X_H]$ is connected.  
  %Property~\tref{ancestor-edge} implies that, for every $xy\in E(H)$, one of $x$ or $y$ is a $T$-ancestor of the other. 
  Choose $a(v)\in X_H$ to be any member of $X_H$ that does not have a strict $T$-ancestor in $X_H$.  Transitivity of the $T$-ancestor relationship, \tref{ancestor-edge}, and connectivity of $H[X_H]$ implies that $a(v)$ is a $T$-ancestor of every node $x\in X_H$, as required.
\end{proof}

For each $x\in V(T)$, we define the \emph{separator} $S_x := \{v\in V(G): a(v)= x\}$.
Observe that $\mathcal{S}:=(S_x : x\in V(T))$ is a partition of $V(G)$.  Let $J:=G^\mathcal{P}/\mathcal{S}$ denote the resulting quotient graph and let $V(J)\subseteq V(T)$ in the obvious way, so that each $x\in V(J)$ is the vertex obtained by contracting $S_x$ in $G^{\mathcal{P}}$. (Nodes $x\in V(T)$ with $S_x=\emptyset$ do not contribute a vertex to $J$.)

@@

%%%%%%%%%%%%%%%%%%%
\section{\boldmath $k$-Planar Graphs: Proof of \cref{k-planar}}
\label{sec-k-planar}

@@

We can reduce the layered width of the $H$-partition of $G$ from $O(k^3)$ to $O(k^2)$ by observing that the dummy vertices in $V(G_0)\setminus V(G)$ do not contribute to the layered width of this partition.  In this setting, the proof of \cref{clm:general-width} is simpler since each vertex $w\in Y_x$ contributes at most two vertices to $V_i\cap Y_x$.  More precisely, each path $P\in\mathcal{P}$ containing an internal (dummy) vertex $w\in Y_x\cap (V_{i-j}\cup V_{i+j})$ contributes: (i)~at most two vertices to $S_x\cap V_i$ for $j\in\{0,\ldots,\floor{(k+1)/2}\}$; (ii)~at most one vertex to $S_x\cap V_j$ for $j\in\{\floor{(k+1)/2}+1,\ldots,k+1\}$; or (iii)~no vertices to $S_x\cap V_j$ for $j > k+1$.
Redoing the calculation at the end of the proof of \cref{clm:general-width} then yields $ |S_x\cap Y_i| \le d\ell\left( 
       2 
       + 4\left\lfloor\frac{k+1}{2}\right\rfloor 
       + 2\left\lceil\frac{k+1}{2}\right\rceil
       \right)
        =
       d\ell\left(
         2 + 2(k+1) + 2\left\lfloor\frac{k+1}{2}\right\rfloor
       \right) \\
        \le
       d\ell(3k+5) 
      = 18k+30$.

With this change, the layered width of the partition given by \cref{modern-major-general} becomes $(18k+30)(k+1)=18k^2+48k+30$.  This establishes \cref{k-planar}.
% Note to self-and-others. In our original SODA submission we had 18k^2+30k here because we forgot that the layer compression step has compress k+1 consecutive layers, not just k consecutive layers.
 Exactly the same approach using \cref{SurfacePartition}(b) instead of \cref{PlanarPartition}(b) establishes 
\cref{g-k-planar} for $(g,k)$-planar graphs.

%%%%%%%%%%%%%%%%%%%%%%%%
\section{Applications}
\label{Applications}
Our main theorems have a number of applications including queue layouts, non-repetitive colourings and clustered colourings  of $k$-planar and $(g,k)$-planar graphs. We present here queue layout results and defer non-repetitive colourings and clustered colourings to \cref{applmore}.

\begin{lem}[\citep{dujmovic.joret.ea:planar}]
\label{qn}
If a graph $G$ has an $H$-partition of layered width $\ell$, then  $\qn(G)\le 3 \ell \, \qn(H) + \floor{\tfrac{3}{2}\ell}$.
\end{lem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Graph Classes: Examples}
\label{Examples}
There are several examples of graph classes that can be obtained from a shortcut system applied to graphs of bounded Euler genus.  These include map graphs, string graphs, powers of bounded degree graphs and $k$-nearest-neighbour graphs.  We present here  $k$-nearest-neighbour graphs results and defer the others to \cref{exmore}.

\bibliographystyle{plainnat}
\bibliography{k-planar}

% % NOTE THAT MANUAL CHANGES HAVE BEEN MADE BELOW.
% \fontsize{11}{11.75}\selectfont 
% \begin{thebibliography}{33}
% \providecommand{\natexlab}[1]{#1}
% \providecommand{\url}[1]{\texttt{#1}}
% \expandafter\ifx\csname urlstyle\endcsname\relax
%   \providecommand{\doi}[1]{doi: #1}\else
%   \providecommand{\doi}{doi: \begingroup \urlstyle{rm}\Url}\fi
% \bibitem[{\'{A}}brego et~al.(2011){\'{A}}brego, Monroy,
%   Fern{\'{a}}ndez{-}Merchant, Flores{-}Pe{\~{n}}aloza, Hurtado,
%   Sacrist{\'{a}}n, and Saumell]{abrego.munroy.ea:on}
% Bernardo~M. {\'{A}}brego, Ruy~Fabila Monroy, Silvia Fern{\'{a}}ndez{-}Merchant,
%   David Flores{-}Pe{\~{n}}aloza, Ferran Hurtado, Vera Sacrist{\'{a}}n, and
%   Maria Saumell.
% \newblock On crossing numbers of geometric proximity graphs.
% \newblock \emph{Comput. Geom.}, 44\penalty0 (4):\penalty0 216--233, 2011.
% \newblock \doi{10.1016/j.comgeo.2010.11.003}.
% \bibitem[Ajtai et~al.(1982)Ajtai, Chv{\'a}tal, Newborn, and
%   Szemer{\'e}di]{ajtai.chvatal.ea:crossing-free}
% Mikl{\'o}s Ajtai, Va{\v{s}}ek Chv{\'a}tal, Monroe~M. Newborn, Endre
%   Szemer{\'e}di.
% \newblock Crossing-free subgraphs.
% \newblock \emph{Ann. Discrete Math.}, 12:\penalty0 9--12, 1982.
% \newblock \doi{10.1016/S0304-0208(08)73484-4}.
% \bibitem[Alam et~al.(2018)Alam, Bekos, Gronemann, Kaufmann, and
%   Pupyrev]{ABGKP18}
% Jawaherul~Md. Alam, Michael~A. Bekos, Martin Gronemann, Michael Kaufmann, Sergey Pupyrev.
% \newblock Queue layouts of planar 3-trees.
% \newblock In Therese~C. Biedl and Andreas Kerren, editors, \emph{Proc. 26th
%   International Symposium on Graph Drawing and Network Visualization} (GD '18), vol.~11282 of \emph{Lecture Notes in Computer Science}, pp.~213--226. Springer, 2018.
% \newblock \doi{10.1007/978-3-030-04414-5\_15}.
% \bibitem[Alon et~al.(2002)Alon, Grytczuk, Ha{\l}uszczak, and
%   Riordan]{AGHR-RSA02}
% Noga Alon, {Jaros{\l}aw} Grytczuk, Mariusz Ha{\l}uszczak, Oliver Riordan.
% \newblock Nonrepetitive colorings of graphs.
% \newblock \emph{Random Structures Algorithms}, 21\penalty0 (3-4):\penalty0
%   336--346, 2002.
% \newblock \doi{10.1002/rsa.10057}.
% \bibitem[Bodlaender(1998)]{Bodlaender-TCS98}
% Hans~L. Bodlaender.
% \newblock A partial $k$-arboretum of graphs with bounded treewidth.
% \newblock \emph{Theoret. Comput. Sci.}, 209\penalty0 (1-2):\penalty0 1--45,
%   1998.
% \newblock \doi{10.1016/S0304-3975(97)00228-4}.
% \bibitem[Bonamy et~al.(2019)Bonamy, Gavoille, and
%   Pilipczuk]{bonamy.gavoille.ea:shorter}
% Marthe Bonamy, Cyril Gavoille, and Micha{\l} Pilipczuk.
% \newblock Shorter implicit representation for planar graphs.
% \newblock In S.~Cabello and B.~Mohar, editors, \emph{9th Slovenian
%   International Conference on Graph Theory} (SICGT), p.~63, June 2019.
% \bibitem[Bose et~al.(2001)Bose, Morin, Stojmenovi\'c, and
%   Urrutia]{bose.morin.ea:routing}
% Prosenjit Bose, Pat Morin, Ivan Stojmenovi\'c, Jorge Urrutia.
% \newblock Routing with guaranteed delivery in ad hoc wireless networks.
% \newblock \emph{Wireless Networks}, 7\penalty0 (6):\penalty0 609--616, 2001.
% \newblock \doi{10.1023/A:1012319418150}.
% \bibitem[Bose et~al.(2012)Bose, Dujmovi\'c, Hurtado, Iacono, Langerman, Meijer,
%   Sacrist\'an, Saumell, and Wood]{ProximityGraphs}
% Prosenjit Bose, Vida Dujmovi\'c, Ferran Hurtado, John Iacono, Stefan Langerman,
%   Henk Meijer, Vera Sacrist\'an, Maria Saumell, David~R. Wood.
% \newblock Proximity graphs: $e$, $\delta$, $\delta$, $\delta$, $\chi$ and
%   $\omega$.
